---
title: "Housing Price vs. Supply 2024"
author: Stephanie Armstrong, Ilgaz Kuscu, Alexander D. Silberman, Baylee Wechsler
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

## Todo:

- finish describing variables

```{r}
rm(list=ls())
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      root.dir = "C:\\Users\\alexa\\Code\\GW DATS\\6101 Intro to Data Science\\Project 1\\GitHub",
                      warning = F, message = F)
options(scientific = T, digits = 3)
```

## Load requisite packages and data

```{r libraries}
library(ezids)
library(ggplot2)
library(dplyr)
```

This project uses the Home Price Appreciation Index and Months' Remaining Inventory dataset from the American Enterprise Institute Housing Center, found [here](https://www.aei.org/home-price-appreciation-index-and-months-remaining-inventory/). The variables are:  

Variable |  Definition  
  :-:    |  :-- 
metro	 | Metropolitan area
price_tier | Category of price data
time_period | Date when the data was collected
hpa_cumul | 
hpa_yoy | 
months_supply | 
decile_hpa_cumul | 
decile_hpa_yoy | 
decile_months_supply | 
low_vs_low_med | 
lo_med_vs_med_high | 
med_high_vs_high | 

```{r read_data}
housing = read.csv("..\\Data\\Copy-of-hpa_supply_download_2025m7.csv")
housing %>% slice_sample(n=5)
```

The data is limited to the year 2024 and cleaned of NA values.

```{r clean_data}
housing_2024 = housing %>% slice(grep("2024",housing$time_period)) %>% na.omit
housing_2024$metro = as.factor(housing_2024$metro)
housing_2024$price_tier = as.factor(housing_2024$price_tier)
housing_2024 %>% slice_sample(n=5)
```

Take a look at the data

```{r}
xkablesummary(housing_2024)
```


