---
title: "Housing Price vs. Supply 2024"
author: Stephanie Armstrong, Ilgaz Kuscu, Alexander D. Silberman, Baylee Wechsler
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

## Todo:

- answer research question

```{r}
rm(list=ls())
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      root.dir = "C:\\Users\\alexa\\Code\\GW DATS\\6101 Intro to Data Science\\Project 1\\GitHub",
                      warning = F, message = F)
options(scientific = T, digits = 3)
```

## Load requisite packages and data

```{r libraries}
library(ezids)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyverse)
library (tidyr)
library(janitor)
library(scales)
library(ggrepel)
library(corrplot)
```

This project uses the State and County Housing Market Indicators dataset from the American Enterprise Institute Housing Center, found [here](https://www.aei.org/the-state-of-the-housing-market/). The variables are:  

Original Variable Name | New Variable Name | Definition  
  :-: |  :-: | :-- 
State	| State |state
County | County_Name |County
FIPS | FIPS_County_Code | 5-digit Federal Information Processing Series codes (first 2 digits indicate state, last 3 indicate sub-county entity)
Year | Year | Year when the data was collected
Tier | Affordability | Categorizes home sales into entry-level (<=80th percentile of FHA sales prices), move-up (all others), and all
Median.Sale.Price..in.Thousands. | Median_Sale_Price_in_k | Median sale price in thousands of USD per county
House.Price.Appreciation.since.2012 | House_Price_Appreciation_since_2012_percent | Cumulative home price appreciation since 2012
House.Price.Appreciation..Year.over.Year | House_Price_Appreciation_yr_over_yr_percent |Home price appreciation since the previous year
Months..Supply | Months_Supply | Number of months it would take for the inventory of existing homes for sale to be exhausted at the current sales pace
New.Construction.Share.of.Sales | New_Constr_by_share_of_sales_percent | Percent of sales comprising new construction
Mortgage.Default.Rate | Mortgage_Default_Rate_percent | AEI Mortgage Default Rate, a measure of how loans originating in a given month would perform under the same conditions as the 2007 financial crisis (<=7%: Low Risk; between 7.01% and 14%: Medium Risk; >14%: High Risk)

```{r read_data}
housing = read.csv("..\\Data\\state_county_data_download_2025.csv")
housing %>% slice_sample(n=5)
```

The data is limited to the year 2024 and cleaned of NA values, and the variables are renamed for clarity.

```{r clean_data}
housing_2024 = housing %>% filter(housing$Year == 2024) %>% na.omit %>%

#rename cols
rename(
  Median_Sale_Price_per_k = Median.Sale.Price..in.Thousands.,
  House_Price_Appreciation_yr_over_yr_percent = House.Price.Appreciation..Year.over.Year.,
  House_Price_Appreciation_since_2012_percent = House.Price.Appreciation.since.2012,
  Months_Supply = Months..Supply,
  New_Constr_by_share_of_sales_percent = New.Construction.Share.of.Sales,
  Mortgage_Default_Rate_percent = Mortgage.Default.Rate,
  County_Name = County,
  FIPS_County_Code = FIPS, 
  Affordability = Tier
)
head(housing_2024)
```

The typing of the variables is also corrected. Some require the symbols "$" and "%" to be removed beforehand, so that is also done.


```{r correct_var_typing}
# as factors
housing_2024$State = as.factor(housing_2024$State)
housing_2024$County_Name = as.factor(housing_2024$County_Name)
housing_2024$FIPS_County_Code = as.factor(housing_2024$FIPS_County_Code)
housing_2024$Affordability = as.factor(housing_2024$Affordability)

# remove prefixes '$' and '%' from values
housing_2024 = housing_2024 %>%
  mutate(Median_Sale_Price_per_k = gsub("\\$", "", Median_Sale_Price_per_k),
         House_Price_Appreciation_since_2012_percent =
           gsub("%","",House_Price_Appreciation_since_2012_percent),
         House_Price_Appreciation_yr_over_yr_percent =
           gsub("%","",House_Price_Appreciation_yr_over_yr_percent),
         New_Constr_by_share_of_sales_percent = gsub("%","",New_Constr_by_share_of_sales_percent),
         Mortgage_Default_Rate_percent = gsub("%","",Mortgage_Default_Rate_percent)
  )

# as num instead of chr
housing_2024$Median_Sale_Price_per_k = as.numeric(housing_2024$Median_Sale_Price_per_k)
housing_2024$House_Price_Appreciation_since_2012_percent =
  as.numeric(housing_2024$House_Price_Appreciation_since_2012_percent)
housing_2024$House_Price_Appreciation_yr_over_yr_percent =
  as.numeric(housing_2024$House_Price_Appreciation_yr_over_yr_percent)
housing_2024$New_Constr_by_share_of_sales_percent = 
  as.numeric(housing_2024$New_Constr_by_share_of_sales_percent)
housing_2024$Mortgage_Default_Rate_percent = as.numeric(housing_2024$Mortgage_Default_Rate_percent)

# For some reason is rounding the data in quite a weird way—inaccurately

# view data
housing_2024 %>% slice_sample(n=5)
```

```{r tier_def}
# def= 
#   {
#     Low: all sales below the 40th percentile of FHA sales prices
#     Low-medium: all sales at or below the 80th percentile of FHA sales prices
#     Medium-high: all sales at or below 125% of the GSE loan limit
#     High: all other sales
#     entry-level: low and low-medium price tiers
#     move-up: medium-high and high price tiers
#   }
```

## Exploratory Data Analysis (EDA)

```{r}
xkablesummary(housing_2024)
```

### boxplot of median sale price by state

```{r boxplot_median_sale_price_by_state}
ggplot(housing_2024, aes(x = reorder(State, -Median_Sale_Price_per_k, median), 
                         y = Median_Sale_Price_per_k)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Distribution of Median Sale Prices by State (2024)",
    x = "State",
    y = "Median Sale Price (in thousands)"
  ) +
  theme_minimal()
```

There are too many to be particularly useful. You can see general trends, but I am going to run this with a smaller state sample.

### boxplot of top and bottom 5 states by housing count

```{r boxplot_top_bottom_5_by_count}
top_states <- housing_2024 %>%
  dplyr::count(State, sort = TRUE) %>%
  dplyr::slice_max(n, n = 5)

bottom_states <- housing_2024 %>%
  dplyr::count(State, sort = TRUE) %>%
  dplyr::slice_min(n, n = 5)

# merge top and bottom states
housing_compare <- housing_2024 %>%
  filter(State %in% c(top_states$State, bottom_states$State)) %>%
  mutate(StateGroup = case_when(
    State %in% top_states$State ~ "States with the Most Houses",
    State %in% bottom_states$State ~ "States with the Fewest Houses"
  ))%>%
  mutate(StateGroup = factor(StateGroup, levels = c(
    "States with the Most Houses",
    "States with the Fewest Houses"
  )))

#plot top bottom comparison ####
ggplot(housing_compare, aes(x = State, y = Median_Sale_Price_per_k, fill = StateGroup)) +
  geom_boxplot() +
  facet_wrap(~ StateGroup, scales = "free_x") +
  labs(
    title = "Median Sale Price in States with Most vs Least Housing Records",
    subtitle = "The Median Sale Price in States with a Larger Supply of Houses is Significantly Lower\nthan States with a Smaller Supply of Houses",
    x = "State",
    y = "Median Sale Price (in thousands)"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("States with the Most Houses" = "skyblue", "States with the Fewest Houses" = "red"))
```

### hist of median sale price

```{r hist_of_med_price}
ggplot(housing_2024, aes(x = Median_Sale_Price_per_k)) +
  geom_histogram(binwidth = 50, fill = "skyblue", color = "black") +
  scale_x_continuous(labels = scales::dollar_format(prefix = "$", suffix = "k")) +
  labs(
    title = "Distribution of Median Sale Prices (2024)",
    x = "Median Sale Price (in thousands)",
    y = "Count"
  ) +
  theme_minimal()
#needs a subtitle with commentary, just remembered all need a caption with the source named, and we can prob add other elements to this, because otherwise it seems to obvi to me
```

### boxplot of sale price by affordability tier

```{r boxplot_price_by_affordability}
ggplot(housing_2024, aes(x = Affordability, y = Median_Sale_Price_per_k, fill = Affordability)) +
  geom_boxplot() +
  labs(
    title = "Median Sale Price by Affordability Tier (2024)",
    x = "Affordability Tier",
    y = "Median Sale Price (in thousands)"
  ) +
  theme_minimal()
#no duh, there are more unpurchased expensive houses because people can't afford it 
#not sure how useful this is, but maybe as a starting baseline 
```

### scatterplot of months supply vs median price

```{r scatter_supply_vs_median_price}
ggplot(housing_2024, aes(x = Months_Supply, y = Median_Sale_Price_per_k)) +
  geom_point(aes(color = Affordability), alpha = 0.7) +
  # geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Housing Supply vs Median Sale Price",
    x = "Months of Supply",
    y = "Median Sale Price (in thousands)"
  ) +
  theme_minimal()
#this is a good one, needs a commentary subtitle and some cleaning
# Not at all what I would have expected. I wonder why this is?
```

### correlation heatmap of numeric variables

```{r corr_heatmap_of_numerics}
housing_numeric <- housing_2024 %>%
  select(where(is.numeric)) %>%
  drop_na()

cor_matrix <- cor(housing_numeric)

corrplot(cor_matrix, method = "color", type = "upper", tl.cex = 0.8, addCoef.col = "black", number.cex=0.5)
#interesting to see positive and inverse relationships between variables 
#did not scan for no relationships
```

### compare color scale\

```{r compare_color_scale}
# Create a vector of colors for top states (blue shades) and bottom states (red shades)
top_states_colors <- scales::seq_gradient_pal("lightblue", "navy")(seq(0, 1, length.out = length(unique(housing_compare$State[housing_compare$StateGroup == "States with the Most Houses"]))))
names(top_states_colors) <- unique(housing_compare$State[housing_compare$StateGroup == "States with the Most Houses"])

bottom_states_colors <- scales::seq_gradient_pal("lightpink", "darkred")(seq(0, 1, length.out = length(unique(housing_compare$State[housing_compare$StateGroup == "States with the Fewest Houses"]))))
names(bottom_states_colors) <- unique(housing_compare$State[housing_compare$StateGroup == "States with the Fewest Houses"])

#last point in line for state label 
label_points <- housing_compare %>%
  group_by(State) %>%
  filter(Months_Supply == max(Months_Supply, na.rm = TRUE)) %>%
  ungroup()

# Combine into one color vector
state_colors <- c(top_states_colors, bottom_states_colors)

#all in gray with faceted compare in color with states labeled ####
ggplot() +
  geom_point(data = housing_2024, aes(x = Months_Supply, y = Median_Sale_Price_per_k),
             color = "gray70", alpha = 0.3, size = 2) +
  # geom_path(data = housing_compare,
  #           aes(x = Months_Supply, y = Median_Sale_Price_per_k, group = State, color = State),
  #           size = 1, alpha = 0.8) +
  geom_point(data = housing_compare,
             aes(x = Months_Supply, y = Median_Sale_Price_per_k, color = State),
             size = 2, alpha = 0.5) +
  geom_text_repel(data = label_points,
                  aes(x = Months_Supply, y = Median_Sale_Price_per_k, 
                      label = State,color=State),
                  size = 3.5, stroke=0.01, show.legend = FALSE) +
  facet_wrap(~ StateGroup) +
  scale_color_manual(values = state_colors) +
  labs(
    title = "Housing Supply vs Median Price by State with Grouped Colors",
    subtitle = "Some commentary here",
    x = "Months of Supply",
    y = "Median Sale Price (in thousands)",
    color = "State"
  ) +
  theme_minimal()
#can't read state names, maybe find a way to make it stand out 
```


### both together in gray with compare in color with states labeled
```{r both_together}
ggplot() +
  geom_point(data = housing_2024, 
             aes(x = Months_Supply, y = Median_Sale_Price_per_k), 
             color = "gray70", alpha = 0.3, size = 1) +
  # geom_path(data = housing_compare %>% filter(StateGroup == "States with the Most Houses"),
  #           aes(x = Months_Supply, y = Median_Sale_Price_per_k, group = State, color = State),
  #           alpha = 0.8, size = 1) +
  geom_point(data = housing_compare %>% filter(StateGroup == "States with the Most Houses"),
             aes(x = Months_Supply, y = Median_Sale_Price_per_k, color = State),
             alpha = 0.5, size = 2) +
  # geom_path(data = housing_compare %>% filter(StateGroup == "States with the Fewest Houses"),
  #           aes(x = Months_Supply, y = Median_Sale_Price_per_k, group = State, color = State),
  #           alpha = 0.8, size = 1) +
  geom_point(data = housing_compare %>% filter(StateGroup == "States with the Fewest Houses"),
             aes(x = Months_Supply, y = Median_Sale_Price_per_k, color = State),
             alpha = 0.5, size = 2) +
  scale_color_manual(values = state_colors) +
  labs(
    title = "Housing Supply vs Median Price: All Counties with Highlights",
    subtitle = "Gray points: All counties | Blue shades: States with Most Houses | Red shades: States with Fewest Houses",
    x = "Months of Supply",
    y = "Median Sale Price (in thousands)",
    color = "State"
  ) +
  theme_minimal()
```

